Flat-histogram method comparison on the 2D Ising model

### 2020-05-08 ising/ising-body.tex
fix some 1/t and edit abstract and stuff

Because of the uncertainty of convergence for WL, many studies have been undertaken
to understand how the modification (or update) factor $\gamma$ impacts the
convergence~\cite{zhou2005understanding,lee2006convergence,
belardinelli2007wang}. Belardinelli and Pereyra showed that an update factor that decreases faster than $1/t$ leads to nonconvergence~\cite{belardinelli2007wang, belardinelli2008analysis, zhou2008optimal}, where $t$ corresponds to the number of moves. Schneider \emph{et al.} outline minor refinements algorithm including scaling the update factor with the number of energy bins~\cite{schneider2017convergence}.
These studies led to the
formation of the $1/t$-WL algorithm and also sparked some debate as to whether
the entire simulation should be split into multiple parts, called ``production run" WL, in order to preserve ergodicity and
detailed balance~\cite{jayasri2005wang, mukhopadhyay2008monte}.

@@

In this work, we compare the convergence properties of five flat-histogram
methods. We detail how each method is implemented and apply the family of
weight-based flat-histogram Monte Carlo methods (pure WL, WL followed by a
"production run", 1/t-WL, SAMC, and SAD) to the 2D Ising model.

@@

\section{Conclusion}
We find that SAD, 1/t-WL, and WL followed by a production run
(with an adequately small $\gamma^{\min}$) demonstrate excellent and robust
convergence. They all converge more rapidly than SAMC, and unlike pure WL do
not suffer from
error saturation. We find that for larger Ising systems SAD reduces the update
factor more slowly (and conservatively) than 1/t-WL and WL followed by a
production run. This means that SAD will take
proportionately more moves to converge to the same value as 1/t-WL as system
size is increased (WL methods are given the energy range \emph{a priori.} and
converge independent of system size). SAD requires the user to specify a
temperature range of interest rather than an energy range of interest as the WL
methods do. The 2D
Ising model is a system where the energy range is known \emph{a priori.}, which
makes a comparison between WL methods and SAD (which requires a minimum
temperature) more biased towards WL methods. For the majority of cases in which
the energy range of the system is not known and where a range of desired
temperatures is known, this will make the SAD method
considerably more convenient. This work also visually demonstrates
that WL followed by a production run performs extremely well and is preferable
to pure WL for ensuring both ergodicity and detailed balance.

>>>>

Because of the uncertainty of convergence for WL, many studies have been undertaken
to understand how the modification (or update) factor $\gamma$ impacts the
convergence~\cite{zhou2005understanding,lee2006convergence,
belardinelli2007wang}. Belardinelli and Pereyra showed that an update factor that decreases faster than $1/t$ leads to nonconvergence~\cite{belardinelli2007wang, belardinelli2008analysis, zhou2008optimal}, where $t$ corresponds to the number of moves. Schneider \emph{et al.} outline minor refinements algorithm including scaling the update factor with the number of energy bins~\cite{schneider2017convergence}.
These studies led to the
formation of the $1/t$-WL algorithm and have also led some researchers to follow
WL with a ``production run'' with fixed weights, in order to preserve ergodicity and
detailed balance~\cite{jayasri2005wang, mukhopadhyay2008monte}.
%
To our knowledge, this work is the first test of the convergence
properties of WL followed by a production run with comparison to other
methods.

@@

In this work, we compare the convergence properties of five flat-histogram
methods. We detail how each method is implemented and apply the family of
weight-based flat-histogram Monte Carlo methods (pure WL, WL followed by a
``production run'', $1/t$-WL, SAMC, and SAD) to the 2D Ising model.

@@

\section{Conclusion}
We find that SAD, $1/t$-WL, and WL followed by a production run (with an
adequately small $\gamma^{\min}$) demonstrate excellent and robust
convergence. They all converge more rapidly than SAMC, and unlike pure
WL do not suffer from error saturation. We find that for larger Ising
systems SAD reduces the update factor more slowly (and conservatively)
than $1/t$-WL and WL followed by a production run. This means that SAD
will take proportionately more moves to converge to the same value as
$1/t$-WL as system size is increased.  However, the WL methods are
solving a simpler problem because they are given the energy range
\emph{a priori}, rather than a temperature range of interest such as
SAD requires.  The 2D Ising model is a system where the energy range
is known \emph{a priori.}, which makes a comparison between WL methods
and SAD more biased in favor of WL methods. For the general case in
which the energy range of the system is not known and where a range of
desired temperatures is known, this will make the SAD method
considerably more convenient, and quite possibly more efficient than a
process involving multiple simulations to determine the energy range
of interest. This work also dramatically demonstrates that WL followed
by a production run performs extremely well and is preferable to pure
WL for ensuring both ergodicity and detailed balance.

### 2020-05-08 ising/ising.tex
fix some 1/t and edit abstract and stuff

\begin{abstract}
We examine the convergence of the stochastic approximation with a dynamic
update factor (SAD) algorithm applied to the 2D Ising model. Comparison with
stochastic approximation Monte Carlo (SAMC) and Wang-Landau (WL) methods show that SAD performs robustly and without user input knowledge of an energy range. We also compare SAD with ``production'' run WL with fixed weights which is the first of its kind recorded. Since SAD is more powerful in the common case in which the range of energies is not known in advance, the 2D Ising model presents a unique challenge since the energy range is readily known and easily discovered by WL methods.
\end{abstract}

>>>>

\begin{abstract}
  We compare the convergence of several flat-histogram methods applied
  to the 2D Ising model, including the recently introduced stochastic
  approximation with a dynamic update factor (SAD) method.  We compare
  this method with the Wang-Landau (WL) method, the $1/t$ variant of
  the WL method, and standard stochastic approximation Monte Carlo
  (SAMC).  In addition, we consider a procedure WL followed by a
  ``production run'' with fixed weights that refines the estimation of
  the entropy.  To our knowledge, this work is the first to test this
  approach against other methods.
  %% To our knowledge, the
  %% convergence rate of this production-run WL approach has never been 
%% We examine the convergence of stochastic approximation with a dynamic
%% update factor (SAD) with other flat-histogram methods applied to the 2D
%% Ising model. The comparison with stochastic approximation Monte Carlo (SAMC)
%% and Wang-Landau (WL) methods shows that SAD performs robustly and without user
%% input knowledge of an energy range. We also compare SAD with WL followed by a
%% ``production run'' with fixed weights which is the first comparison of its kind
%% recorded.
We find that WL followed by a production run \emph{does} converge to
the true density of states, in contrast to pure WL.  Three of the
methods converge robustly: SAD, $1/t$-WL, and WL followed by a
production run.  Of these, SAD does not require \emph{a priori}
knowledge of the energy range.
\end{abstract}
